---
title: "771 Class 1"
---


```{r global_options, include=FALSE}
knitr::opts_chunk$set(fig.width=6, fig.height=6, fig.path='Figs/',
                      echo=TRUE, warning=FALSE, message=FALSE)
```


### Wednesday  2023-09-06

Welcome to Stat 771, Computational Statistics.

### Semester plan.

Introductions,  student work, assessment, learning outcomes, textbooks, syllabus and other course information....what we're going to try to do this semester!

  
We talked about various computation systems and that we're demonstrating things in R and 
RStudio this term.
  
Next we did a short overview of examples of issues with finite-precision arithmetic.

### Being careful with numbers.

E.g. Numbers in R


```{r}
options(digits=8)
10
```

```{r}
1/10
```

R is good with infinities

```{r}
1/0
```

```{r}
1+Inf
```

```{r}
Inf-Inf
```

But R uses finite-precision arithmetic, and is necessarily limited in how it represents numbers. For example, small positive numbers have to be larger than a fixed lower limit.

```{r}
1/1000000 > 0

1e-100 > 0

1e-300 > 0

1e-400 > 0
```

It may seem that such a small number won't come up in routine statistics, but there are many situations where these do (e.g. likelihood calculations).

R is also limited in the size of big numbers.

```{r}
10^10 < Inf
10^300 < Inf
10^400 < Inf
```
The 64-bit precision confers about 18 decimal digits, and so there is imprecision beyond that. For example

```{r}
options(digits=20)
pi
factorial(5)
factorial(23)
```
The first and last of these clearly differ from the mathematical objects.

These precision limitations affect routine computations, and so we need to keep them in mind. For example, the following computes the likelihood for the mean parameter $\mu$ of a normal model (with known variance 1), for a random sample.

```{r}
L <- function(mu,dat)
 {
  tmp <- dnorm(dat, mean=mu)
  bar <- prod(tmp)
  return(bar)
}



# an example data set
data(PlantGrowth)
tmp <- PlantGrowth[,1] 
x1 <- tmp[PlantGrowth[,2]=="ctrl"]


L(4, dat=x1)
L(5, dat=x1)



```

To look at the whole likelihood function for either data set, let's work on a grid

```{r}
mu.grid1 <- seq(3,7,length=100)
```

and vectorize the likelihood evaluator to handle all grid values at once

```{r}
Lv <- Vectorize(L,vectorize.args="mu")

lik1 <- Lv(mu.grid1, dat=x1)
```

All's fine in the first data example

```{r}

plot( mu.grid1, lik1, type="l")
abline( v=mean(x1), col="red") ## the MLE
rug(x1,col="blue")
```

but underflow has killed the second case

```{r}
data(treering)
x2 <- treering

mu.grid2 <- seq(0,2,length=100)
lik2 <- Lv(mu.grid2, dat=x2)


plot( mu.grid2, lik2, type="l" )
rug(x2,col="blue")
```
We avoid the underflow by working with log likelihoods.

```{r}
l <- function(mu,dat)
 {
  tmp <- dnorm( dat, mean=mu, log=TRUE )
  bar <- sum( tmp )
  return(bar)
 }

lv <- Vectorize(l, vectorize.args="mu" )

loglik1 <- lv( mu.grid1, dat=x1 )
loglik2 <- lv( mu.grid2, dat=x2 )

plot( mu.grid1, loglik1, type="l")
abline( v=mean(x1), col="red")
rug(x1,col="blue")

plot( mu.grid2, loglik2, type="l")
abline( v=mean(x2), col="red")
rug(x2,col="blue")

```

Here we're using raw computing to get at what math would tell us is the sample mean.
Underflow affected the previous example, and we see the scale of log probability densities (negative thousands in example 2) gets us well below log(1e-400) investigated above.   We can run into related problems even when the numbers involved are not near the limit of precision.  For example

```{r}
options(digits=8)
v <- c(1, 1e-20)
v[1]
v[2]
v[1]+v[2] > v[1]
```
The problem here is that the mathematical value requires more significant digits than are available in base R.

When working with small positive numbers, say $a_1, a_2, \cdots, a_n$, we often use the logarithmic scale to record intermediate computations.   For example, suppose we wish to resolve $S = \sum_{i=1}^n a_i$, and since we know it's the sum of small numbers and likely to be small, we will be happy to compute $\log(S)$.  Further, we are working with $a_i$'s on the log scale, and have in the computer $b_i = \log(a_i)$.  Naively, (for a test set of logged numbers in the vector b), we could do

```{r echo=TRUE}
b <- dnorm( x1, mean=50, log=TRUE ) # example
ans <- log(sum(exp(b)))
ans
```
but underflow has affected the inner expectation, which is avoided in the following <i> log-sum-exp trick </i>:

```{r echo=TRUE}
m <- max(b)
ans2 <- m + log(sum(exp(b-m)))
ans2
```
The log-sum-exp trick is based on simple math you can verify. Manipulating sums of small numbers often happens in Bayesian and mixture model computations, which we'll learn more about later.

We note that advances have been made in software that tricks a fixed-precision system into having higher precision.  In R, you may like to check out the <i> Rmpfr </i> package to see how this goes.  Here are two examples that correct the imprecision of base R noted above.  

```{r echo=TRUE}
library(Rmpfr)
Pi <- Const("pi", prec=260)
Pi

factorialMpfr(23)
```

These high-precision numbers look better than what we got from the defaults.


A nice thing about <i> Rmpfr </i> is that we can use the high-precision numbers 
 to check the accuracy of various computational schemes. For example, let's check
the accuracy of `log` and the function `log1p` which aims to more accurately compute $\log(1+x)$ when the magnitude of $x$ is very small (and thus were $\log$ is subject to underflow).
  The plot first looks at the relative error of `log`.


```{r echo=TRUE}
library(Rmpfr)

powers<- seq(13,17, length=100)

x <- 1/10^powers

x. <- mpfr(1, precBits=256)/10^powers

a <- log( 1+x )
aa <- log1p( x )

a.  <- log( 1+ x. )
aa. <- log1p( x. )
## a. and aa. are very close; either may be considered the `truth`

# relative error

re1 <- (a - a.)/a.
re2 <- (aa - aa.)/aa.

plot( powers, re1 , main="error in log(1+x); x=10^[-power]",
		ylab="relative error" )
lines( powers, re2, col="red", lwd=2 )
legend( "bottomleft", legend=c("log1p(x)"), lwd=2, col="red" )
```

So `log1p` is much more acccurate than `log` for certain arguments.  If we zoom in, we see that `log1p` too is in error, but at a much lower level:


```{r echo=TRUE}
plot( powers, re2, main="error in log1p(x); x = 10^[-power]",
		ylab="relative error" )
abline( h=0 )
```


Here are another curious example using pseudo-random numbers (we'll learn a lot more about
these things going forward)

```{r echo=TRUE}
B <- 10^4
ok <- logical(B)

for( b in 1:B )
 {
  u <- runif(3)
  left <- (u[1]+u[2])*u[3]
  right <- u[1]*u[3] + u[2]*u[3] 
  ok[b] <- left == right
 }
 
print( mean(ok) )

```
It seems that the distributive law isn't always obeyed by finite-precision arithmetic!
Indeed, we know that when checking for equality we really ought to check that differences
are smaller than some pre-specified positive tolerance.

If we use high-precision numbers, the distributive law holds, thankfully!

```{r echo=TRUE}
B <- 1000
ok <- logical(B)
for( b in 1:B )
 {
  u <- runif(3)
  u. <- mpfr(u, precBits=100 )
  left <- (u.[1]+u.[2])*u.[3]
  right <- u.[1]*u.[3] + u.[2]*u.[3] 
  ok[b] <- left == right
 }

print(mean(ok))
```
You may notice that the cost of working in high-precision is computer speed!  It's much
slower to compute (so I reduced `B` in this case).


Here's a final example (not covered in class), which exemplifies a well-known issue with
finite-precision arithmetic that we lose precision when taking differences between numbers
of similar magnitude.



```{r echo=TRUE}
library(Rmpfr)

powers<- seq(13,17, length=100)

mu <- 10^powers
n <- length(powers)

set.seed(7) 
e1 <- rnorm(n)
e2 <- rnorm(n)

x <- mu+e1  ## these are vectors of numbers where each x_i is relatively close to y_i
y <- mu+e2

x. <- mpfr(mu, precBits=512 ) +  mpfr(e1, precBits=512) ## high-precision versions
y. <- mpfr(mu, precBits=512 ) +  mpfr(e2, precBits=512)

a <- x-y     
a. <- x.-y. 

# relative error

re1 <- (a - a.)/abs(a.)

plot( mu, a. , main="true differences x.-y.", log="x" )
```

The plot above shows the `true` (i.e. high-precision) deviations `a=x-y` as a function of
the underlying means.  Below shows computed differences, followed by an overlay plot which
shows for large `mu` that errors start to creep in.

```{r echo=TRUE}
plot( mu, a , main="computed differences x-y", log="x" )
```

```{r echo=TRUE}
plot( mu, a. , main="",  log="x" )
points( mu, a, pch=19, col="red" )
```

And the relative error
```{r echo=TRUE}
plot( powers, re1 , main="error in difference", ylab="relative error" )
```





### some references

* [Knut Martin Mørken's notes on round-off](https://www.uio.no/studier/emner/matnat/math/MAT-INF1100/h08/kompendiet/round-off.pdf)

* [Wikipedia](https://en.wikipedia.org/wiki/Round-off_error)

* What Every Computer Scientist Should Know About Floating-Point Arithmetic
[Goldberg, D, 1991](https://docs.oracle.com/cd/E19957-01/806-3568/ncg_goldberg.html)

* FAQs from CRAN
[See section 7.31](https://cran.r-project.org/)

* R Multiple Precision Floating Point Reliable 
 [Rmpfr](https://cran.r-project.org/web/packages/Rmpfr/index.html)

* Nice R notes from [Phil Spector](https://www.stat.berkeley.edu/~spector/Rcourse.pdf)

* A clear tutorial on R numbers by [Patrick Burns](http://www.burns-stat.com/documents/tutorials/impatient-r/more-r-key-objects/more-r-numbers/)
