---
title: "771 Class 2"
---


```{r global_options, include=FALSE}
knitr::opts_chunk$set(fig.width=6, fig.height=6, fig.path='Figs/',
                      echo=TRUE, warning=FALSE, message=FALSE)
```


### Monday 2023-09-11

## More basics

It's important for us to have some understanding how common statistical procedures
are computed and also what common approaches are available in standard and non-standard
settings. 


### Normal tail areas

We often need to compute tail probabilities in a standard normal distribution.  In
the old days students did this by looking in tables at the back of a book (but someone had to
make the book!); nowadays
we use `pnorm` in `R`, or something comparable in other systems.   We seek to compute,
for inpute $x$, the numerical value

$$ P( Z \leq x ) = \int_{-\infty}^x \phi(y) \, dy$$

where $\phi(y) \propto \exp( -y^2/2 )$ is the standard bell curve.   There's no
closed formula for this probability, sometimes called $\Phi(x)$.    There are many ways to 
compute $\Phi(x)$, for instance:

```{r}

x <- 1.96
p <- 1/2 + integrate(dnorm, lower=0, upper=x )$value
p

```

The built-in `integrate` function deploys a numerical method that we'll investigate later, but
we note `pnorm` does not involve an explicit integration.



### Continued fractions

We recall the golden ratio $g = 1+1/g = (1+\sqrt{5})/2$. This number has interested thinkers
for thousands of years; two line segments in this ratio have the property that the longer over the
shorter equals the longer plus the shorter over the longer.   By repeated substitution:

$$ g = 1/[1+1/[1+1/[1+1/[1+1/[1+...]]]]] $$

We also recall (without proof, but with amazement) that Laplace
established a continued fraction for the Mill's ratio $R(t) = [1-\Phi(t)]/\phi(t)$.

In particular

$$ R(t) = 1/[t+1/[t+2/[t+3/[t+4/[t+5/[t+...]]]]]]   $$

We note that a nice **recursion** forms by truncating this expansion.  For instance
if we truncate at level 5; e.g. with input $t=1.96$,

```{r echo=TRUE}
t <- 1.96

d <- t
d <- t+5/d
d <- t+4/d
d <- t+3/d
d <- t+2/d
d <- t+1/d

millsRatio <- 1/d

pnormapprox <- 1-dnorm(t) * millsRatio

pnormapprox

```
This agrees to four decimal places with `pnorm`.


#### Expand the integrand and integrate term by term

Recalling the expansion  $e^u=1+u+u^2/2+...$, we get, for $x>0$, 

$$ \Phi(x) = 1/2 + \sum_{m=0}^\infty (-1)^m  x^{(2m+1} /( (2^m) (m!) (2m+1) )   $$

By truncating this sum at some big number $B$ we have a ready approximation. 
Though this suffers from being an alternating series, and later terms have large numerators and denominators,  and is thus subject to substantial round-off error.  It's not the basis of `pnorm`.

#### Taylor expand $g(x) = [ \Phi(x) - 1/2 ]/\phi(x) ]$

We check that $g(-x) = -g(x)$ (i.e., $g$ is odd), and then leverage that $g$ must
therefore have an expansion around zero in odd powers of $x$:

$$ g(x) = \sum_{m=0}^\infty c_m x^{2m+1}  $$

We then do a bit of work to identify the coefficients, finding 

$$ c_m = [ (2m+1)!! ]^{-1}  $$

where the double factorial sign means product of odd terms down to 1.   Well now
we have a nice positive term series for $g$, which we can readily truncate to get
an approximation of $g$ and then $\Phi$  

**Note:**  it happens here and in many cases elsewhere that a probability density is
 readily evaluated but corresponding probability masses are very difficult to 
 compute.

This overcomes the errors due to the alternations in the first Taylor expansion, but
it's still not the preferred method.  



#### Chebychev

In fact `pnorm` uses none of the methods reviewed above, but rather uses an
algorithm coded by Cody (1969) that is based upon a special rational-function
Chebyshev approximation.   Cody approximates the so-called *error function*, a simple
transform and close relative of $\Phi$ using a ratio of two finite polynomials with
known (i.e. computed by Cody) coefficients.    His paper claims high accuracy of
the approximation over a wide range of input $x$.


[Rational Chebyshev Approximations for the Error Function, 
By W. J. Cody; Math. Comp. 23 (1969), 631-637](http://www.ams.org/journals/mcom/1969-23-107/S0025-5718-1969-0247736-4/)


It turns out that the favored algorithm for computing `qnorm`, i.e. the quantiles of 
the standard normal, is also based on Chebyshev approximations, though different
polynomials are involved.

[Wichura, M. J. (1988) Algorithm AS 241: The percentage points of the normal distribution. Applied Statistics, 37, 477–484.](http://www.jstor.org/stable/2347330?seq=1#page_scan_tab_contents)

#### some example computations


```{r echo=TRUE}
# Some code to look at two power series and two continued fraction
# approximations to normal cumulative probability.

# taylor expand density then integrate
taylor1 <- function(x,n)
 {
   nn <- 0:n
   u <- (-1)^nn * x^(2*nn+1) / ( factorial(nn) * 2^nn * (2*nn+1) )
   u <- u/sqrt(2*pi)
   u
 }

# a positive term expansion
taylor2 <- function(x,n)
 {
   nn <- 0:n
   kff <- cumprod( seq(1,(2*n+1), by=2) )
   u <-  dnorm(x) * x^(2*nn+1) / kff
   u
 }

# Laplace, 1812
# Shenton, 1954, eq 16
# R(t) = 1/[t+1/[t+2/[t+3/[t+...]]]]
## and R(t) = (1/phi(t))*[ 1- Phi(t) ]

Two <- 2
Pi <- pi

laplace.cf <- function(x,n)
 {
  u <- x
  for( j in n:1 )
   {
    u <-  x + j/u
   }
  u <- 1- (1/u)*exp( -x^2/2 )/sqrt(Two*Pi) 
  return(u)
 }


#
shenton.cf <- function(x,n)
 {
  u <- 2*n+1
  for( j in n:1)
   {
    u <-  2*j-1 + (-1)^(j) * (j) * x^2 /u
   }
  u <- x/u
  u <- u*exp(-(1/2)*x^2 )/sqrt(2*pi)  + 1/2
  u
  }

## an illustration

B <- 30
x <- 3.5


cf1 <- numeric(B)
cf2 <- numeric(B)
for( i in 1:B)
{
   cf1[i] <- laplace.cf( x, n=i)
   cf2[i] <- shenton.cf( x, n=i)
}


v1 <- taylor1(x, B-1)
v2 <- taylor2(x, B-1)
c1 <- cumsum(v1) 
c2 <- cumsum(v2)
y <- pnorm(x)  - 1/2

e1 <- abs( c1-y )
e2 <- abs( c2-y )
f1 <- abs( cf1 - pnorm(x) )
f2 <- abs( cf2 - pnorm(x) )

u <- c(f1,f2,e1,e2)
u[u==0] <- min( u[u>0]) ## helps log-scale plotting

plot( 1:B,  e1,    col="red", type="b", log="y" , 
		xlab="number of iterations", ylab="absolute error",
	main="Approximating pnorm(x)", ylim=range(u), 
       sub=paste( "x=", format(x,digits=2) ) )

lines(e2,  col="blue", type="b" )

legend( "bottomleft", legend=c("alternating series","positive term series","Laplace CF", "Shenton CF"),
		 col=c("red","blue","magenta","green" ), pch=1 )

lines( f1, col="magenta", type="b" )
lines( f2, col="green", type="b" )

```

### Note on approximation

For any of the approximations above, we have a function, say $f_B(x)$
that converges (mathematically)
 to the target $\Phi(x)$ as $B$ diverges.   In typical implementations, $f_B(x)$ is evaluated
by fixed-precision arithmetic, and so the computed value may differ from the idealized value, owing to round-off error, for example, quite apart from any difference between the ideal (i.e. real) value and the  target. Our numerical experiments compared algorithms at a single value of $x$, but some sort of uniform convergence over a wide range of inputs is desired for a general-purpose tool.

With `Rmpfr`, we can investigate numerical accuracy of any given sequence, though there is the question of
what the true normal tail area number ought to be.   We might think to compute it with the high
precision `pnorm` that accompanies `Rmpfr`, but I suspect that this output is simply a high-precision
implementation of Cody's algorithm, and so it gives a high-precision value of $f_B(x)$ for some $B$, rather
than a high-precision value for the target quantity.   

With large enough $B$ all these approximations will converge to the ideal value, especially when evaluated
by high-precision arithmetic.  


```{r}
## look at high precision upper tail probabilities
library(Rmpfr)
 
x <-  1.96
nmax <- 100


## Use positive term series to get a high precision answer
## (since there's no high precision pnorm...)a

pnorm.posterm.highprec <- function(x,n,pb=1000)
 {
   x. <- mpfr(x, precBits=pb)
   nn <- 0:n
   kff <- cumprod( seq(1,(2*n+1), by=2) )
   u <-  dnorm(x.) * x.^(2*nn+1) / kff
   u
 }

p.hp <- 1/2 - cumsum( pnorm.posterm.highprec(x,n=nmax) )
pnorm.val <- pnorm( mpfr(x, precBits=1000), lower.tail=FALSE )  ## Chebyshev is not exact
							## and this shows error

prob.hp <- p.hp[length(p.hp)]
hh <- pnorm.val - prob.hp ##  

err.posterm <- p.hp - prob.hp 
```

This is more precise than needed for any practical purpose, but we report for interest
a high precision calculation of $1-\Phi(1.96)$: 
```{r}
 print(prob.hp, digits=100)
```

The plot below shows the error dropping, as computed by the positive term expansion from above.
(horizontal line at `pnorm` which itself does not operate in high-precision ).

```{r}
plot( 1:(nmax+1), err.posterm, log="y", xlab="iteration", ylab="absolute error", main="P_{h.p.}[ N(0,1) > 1.96 ]" )

abline( h=as.double(hh))
```



### Poisson Binomials

We review the so-called Poisson-Binomial distribution which is the sum of independent but
not identically distributed Bernoulli trials; say $X_i$ has success probability $\theta_i$.  Though we can compute the joint probability of any particular realization of binary trial outcomes

$$P[ X_1=x_1, X_2=x_2, \cdots, X_n=x_n] = \prod_{i=1}^n [ \theta_i^{x_i} (1-\theta_i)^{1-x_i} ] $$

there's no obvious way to compute the distribution of $S_n=\sum_{i=1}^n X_i$.  At least we know
that  for $s \in \{0,1,2, \cdots, n\}$,
$$
P[S_n=s] = \sum_{x \in \mathcal{X}_{n,s} } \prod_{i=1}^n [ \theta_i^{x_i} (1-\theta_i)^{1-x_i} ]
$$
where $\mathcal{X}_{n,s}$ is the set of all length-$n$ binary vectors containing exactly $s$ 1's.
With non-constant $\theta_i$, this summation does not easily simplify, but we can use some tools to
calculate for small $n$, such as:

```{r}
library(rje)

n <- 9
theta <- (1:n)/(n+1)

ps <- powerSet(1:n,n)

supp <- seq(0,n)
pmf <- rep(0, n+1)
pmf[1]  <- prod( 1-theta )  ##  chance of S=0 is chance of all 0's
pmf[n+1]  <- prod( theta )  ##  chance of S=n is chance of all 1's
# otherwise add up over options

for( j in 2:((2^n)-1)  )  ## first set is empty
 {
  sb <- ps[[j]]  # subset of 1's
  s <- length( sb )   ## 
  ii <- rep(0,n)
  ii[sb] <- 1
  pmf[s+1] <- pmf[s+1] + prod( theta^ii * (1-theta)^(1-ii) )
 }


plot( supp, pmf, type="h", main="Poisson binomial",col="blue" )

```

The method above works well for small $n$ but is infeasible in many relevant examples.  Thomas
and Taub (1982) report a beautiful **recursion** for the Poisson binomial case:

[Calculating binomial probabilities when trial probabilities are unequal]](http://www.tandfonline.com/doi/abs/10.1080/00949658208810534)


Rather than brute force enumeration, Thomas and Taub show how $P(S_n=s)$ may be
related simply to $P(S_{n-1}=t)$.   For a given value $s$ , the only way $S_n=s$ is
if either $S_{n-1}=s$ and $X_n=0$ or $S_{n-1}=s-1$ and $X_n=1$. This suggests a
recursion, coded below.  A matrix is constructed, each row of which is a p.m.f. of
a partial sum.  Each entry in a lower row is computed from at most two entries from the prior row,
giving a computationally efficient alternative to complete enumeration.

```{r echo=TRUE}

poisbin <- function(theta)
        {
        # theta is a length k vector of success probabilities
        # for the Bernoulli trials.  
        k <- length(theta)
        aa <- matrix(NA,k,k+1)
        aa[1,1] <- 1-theta[1]
        aa[1,2] <- theta[1]
        for( i in 2:k )
         {      
          aa[i,1] <- aa[i-1,1] * (1-theta[i])
          aa[i,i+1] <- aa[i-1,i] * theta[i]
          aa[i,2:i] <- aa[i-1,1:(i-1)]*theta[i]+aa[i-1,2:i]*(1-theta[i])
         }
        return( aa[k,] )  # prob mass function on 0:k
        }

# for example
k <- 8
theta <- seq( .01, .6, length=k )
pmf <- poisbin(theta)

plot( 0:k, pmf, type="h", lwd=2, col="blue" )
```

Here's a more challenging example from large-scale empirical Bayes testing:

```{r}

library(fdrtool)
data(pvalues)
hist(pvalues)
ff <- fdrtool(pvalues, statistic="pvalue", plot=FALSE )
theta <- ff$lfdr[ff$lfdr < 0.05 ] ## we say t
```

Think of each $p-$value, from a vector of `r length(pvalues)` is testing some null hypothesis. 
In an empirical Bayes approach, we treat the marginal distribution of $p-$values as a mixture of a uniform (null is true) and something shifted toward the origin (when the null is false).  The `fdrtool` function computesthe *local false discovery rate (lfdr)*, which is an empirically derived posterior probability that the null is true conditional on the $p-$value.   The most interesting units have smallest $p-$values; if we report a list say with *lfdr* bounded by $5 \%$, then this list of discoveries contains `r length(theta)` units.  Each reported unit may or may not constitute a type-I error.  The number of type-I errors is the sum of Bernoulli trials, but their success probabilities are not constant (they depend on the $p-$value).   Treating these
calls as independent, the number of *false discoveries* is like $S$ above.


```{r}
theta <- ff$lfdr[ff$lfdr < 0.05 ] ## we say t

pmf.fd <- poisbin(theta)

plot( 0:length(theta), pmf.fd, type="h", lwd=2, col="blue", xlab="# false discoveries", xlim=c(0,30) )
```



